### 2023-06-28 paper/main.tex
add Schultz affil
..
revision1

\author{\IEEEauthorblockN{Gergely Firtha}
    \IEEEauthorblockA{\textit{Dept. of Networked Systems and Services} \\
        \textit{Budapest University of Technologies and Economics}\\
        H-1111 Budapest, Hungary \\
        firtha@hit.bme.hu}
    \and
    \IEEEauthorblockN{Nara Hahn}
    \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
        \textit{name of organization (of Aff.)}\\
        City, Country \\
        email address or ORCID}
    \and
    \IEEEauthorblockN{Frank Schultz}
    \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
        \textit{name of organization (of Aff.)}\\
        City, Country \\
        email address or ORCID}
    \and
    \IEEEauthorblockN{Péter Fiala}
    \IEEEauthorblockA{\textit{Dept. of Networked Systems and Services} \\
        \textit{Budapest University of Technologies and Economics}\\
        H-1111 Budapest, Hungary \\
        fiala@hit.bme.hu}
}

@@

\section{Introduction}
The aim of sound field synthesis is to reproduce a virtual target sound field over an extended listening area using a densely spaced loudspeaker arrangement, known as the secondary source distribution (SSD).
By feeding the loudspeakers with specific driving functions, the superposition of sound fields from each SSD element should ideally match the target sound field in the intended receiving area.
One prominent sound field synthesis method is Wave Field Synthesis (WFS) \cite{Berkhout1993:Acoustic_control_by_WFS, Start1997:phd}.%, which obtains the necessary driving functions from a suitable boundary integral representation of the target sound field.

\section{Theoretical basics}
\subsection{The local propagation vector:}
Consider an arbitrary steady-state free space sound field at an angular frequency $\omega$.
As a standard ansatz in the field of geometrical acoustics the sound field can be written in the general polar from as
%
\begin{equation}
    P(\vx,\omega) = A(\vx) \te^{- \ti \omega \tau(\vx)},
    \label{eq:Pxw}
\end{equation}
%
where $A(\vx)$ and $\tau(\vx)$ are real-valued functions.
This formulation applies to both plane waves and (3D) point sources.
The propagation dynamics of the sound field are governed by its phase function $\tau(\vx)$, termed as the eikonal in the field of ray acoustics.
In the temporal domain, the sound field can be obtained by taking the inverse Fourier transform of \eqref{eq:Pxw}, yielding:
\begin{equation}
    p(\vx,t) = A(\vx) \, \delta \! \left( t - \tau(\vx)\right).
\end{equation}
From the above formulation it is evident that the eikonal $\tau(\vx)$ describes the propagation delay the wavefront takes to arrive at $\vx$.
The eikonal equation is obtained by substituting the ansatz \eqref{eq:Pxw} into the Helmholtz equation, stating that in a source free volume
\begin{equation}
    \left| \nabla \tau(\vx) \right| = \frac{1}{c}
\end{equation}
is fulfilled.
The gradient of the eikonal is termed as the local propagation vector
\begin{equation}
    \hat{\vk}^P(\vx) = \posvec{3}{\hat{k}_x^P(\vx)}{\hat{k}_y^P(\vx)}{\hat{k}_z^P(\vx)} = c \nabla \tau(\vx,\omega).
\end{equation}
In steady-state, the local propagation vector is perpendicular to the wavefront (equiphase positions) at an arbitrary location with unit length pointing towards the local propagation direction \cite{Firtha2016}.
In the temporal domain it points into the local propagation direction at the time instant of the wavefront passby.

@@

\subsection{Conventional 2.5D Wave Field Synthesis}
Consider a smooth convex secondary source distribution located at $\vxo = \posvec{3}{x_0}{y_0}{0}$ consisting of a continuous distribution of 3D point sources, described by the 3D Green's function.
In this geometry the target field inside the area bounded by the SSD is described by the  Kirchhoff approximation of 2.5D Kirchhoff-Helmholtz integral, from which the 2.5D driving functions can be extracted.
For an arbitrary simple target virtual sound field $P(\vxo,\omega)$ the steady-state driving functions read as \cite{Firtha2016}
\begin{equation}
    D(\vxo,\omega) = \underbrace{\sqrt{8\pi \frac{\ti \omega}{c}}}_{ H_{\text{pre}}(\omega) }
    w(\vxo) \sqrt{d_{\mathrm{ref}}(\vxo)}
    P(\vxo,\omega)
    .
    \label{Eq:WFS25D}
\end{equation}
The driving function consists of
\begin{itemize}
    \item a frequency dependent pre-equalization filter $H_{\text{pre}}(\omega)$,
    \item a secondary source selection window
          \begin{equation}
              w(\vxo) = \max{\left( \hat{k}_{\mathrm{n}}^P(\vxo),0\right)},
          \end{equation}
          where $\hat{k}_{\mathrm{n}}^P(\vxo)$ is the normal component of the local propagation vector,
    \item a gain factor $\sqrt{d_{\mathrm{ref}}(\vxo)}$ allowing amplitude correction along a reference curve, depending on the actual virtual field model (c.f. \cite{Firtha2016}),
    \item and the virtual field measured on the SSD.
\end{itemize}
Assuming a simple virtual sound field as given by \eqref{eq:Pxw} the driving functions can be written as
\begin{equation}
    D(\vxo,\omega) = H_{\text{pre}}(\omega)
    \underbrace{ w(\vxo)  \, \sqrt{d_{\mathrm{ref}}(\vxo)} \, A(\vxo) }_{A^D(\vxo)} \te^{- \ti \omega \tau(\vxo)}
    ,
\end{equation}
and in the time domain as
\begin{equation}
    d(\vxo,t) = h_{\text{pre}}(t) \ast_t  A^D(\vxo)\,  \delta\left( t - \tau(\vxo)\right),
    \label{eq:d_wfs_td}
\end{equation}
with $A^D(\vxo)$ being the real valued overall gain factor of the driving function, $\ast_t$ denoting temporal convolution and $h_{\text{pre}}(t)$ is the temporal WFS prefilter impulse response.
For an analytic FIR pre-equalization filter implementation the reader is referred to \cite{Schultz2016}.
In the following due to the associativity of convolution this prefiltering is excluded from the discussion.

@@

\begin{figure*}[h!]
    \begin{center}
        \begin{overpic}[width = 0.75\columnwidth]{figs/ideal_synthesis.png}
            \footnotesize \put(0,2){(a)}
        \end{overpic} \hspace{2cm}
        \begin{overpic}[width = 0.75\columnwidth]{figs/aliased_synthesis.png}
            \footnotesize \put(0,2){(b)}
        \end{overpic}
        \\
        \begin{overpic}[width = 0.75\columnwidth]{figs/ideal_synthesis_spectrum.png}
            \footnotesize \put(0,0){(c)}
        \end{overpic} \hspace{2cm}
        \begin{overpic}[width = 0.75\columnwidth]{figs/aliased_synthesis_spectrum.png}
            \footnotesize \put(0,0){(d)}
        \end{overpic}
    \end{center}
    \caption{fg}
    \label{Fig:aliasing}
\end{figure*}
Conventional WFS theory assumes a continuous secondary source distribution.
In practical applications the SSD is composed of evenly spaced discrete source elements.
As a result, \emph{aliasing wavefronts} emerge from the individual secondary sources, following the intended virtual wavefront, leading to a \emph{spatial aliasing phenomena}.
Although localization does not degrade due to the precedence effect, aliasing is perceived as strong coloration predominantly in the high-frequency region varying with the receiver position and the virtual source position/direction, occurring predominantly above the \emph{aliasing frequency}.

Figure \ref{Fig:aliasing} illustrates spatial aliasing in the temporal domain through the example of the synthesis of an impulsive plane wave, arriving from $\phi_{\mathrm{PW}} = 0^{\circ}$, applying a discrete secondary source distribution in comparement with a quasi-continuous synthesis scenario.
In the current example the SSD is a circular one with the radius of $R_0 = 2~\mathrm{m}$, consisting of $N_0 = 90$ 3D point sources.
The aliased synthesis in the temporal domain is depicted in Figure \ref{Fig:aliasing} (b).

Mathematically, aliasing can be modeled as the discretization of the theoretically continuous driving functions:
Assume that a suitable parametrization $d(s,t)$ of the secondary source is given, e.g. the arc length $s = R_0\, \cos \phi$ in case of a circular SSD (where $\phi$ is the polar angle), or the linear position on an infinite long linear SSD.
The discretization process is given by sampling the driving function by a sampling function, consisting of a series of Dirac-deltas in the actual secondary source positions:
\begin{equation}
    d_s(s,t) = \sum_{\nu = 0}^{N_0} d(s,t) \, \delta(s - \nu \Delta s) \, \Delta s,
    \label{eq:sampling_td}
\end{equation}
where $\Delta s$ is the sampling arc length.

By investigating the sampling process in the frequency-wavenumber domain, the spectrum of the sampled driving function is given by
\begin{equation}
    \tilde{D}_s(k_s,\omega) = \sum_{\nu = -\infty}^{\infty} \tilde{D}\left(k_s- \nu \frac{2\pi}{\Delta s},\omega\right) \Delta s,
    \label{eq:sampling_fd}
\end{equation}
where $t \rightarrow \omega$ and $s \rightarrow k_s$ are Fourier-transform pairs.
Obviously, for a closed SSD contour the continuous spatial Fourier-transform yields a discrete spectrum, consisting of non-zero components only on the multiples of the spatial wavenumber, i.e. at
\begin{equation}
    k_s = \frac{m}{R_0},
\end{equation}
constituting a Fourier series.

@@

Obviously, spatial aliasing can be avoided by analytically spatial-bandlimiting the driving functions to the Nyquist wavenumber.
This can be performed by either spatially filtering the driving functions with an appropriate spatial low-pass filter prior to numerical evaluation, or performing bandlimitation directly in the wavenumber (modal) domain analytically.
The latter approach is implemented by \cite{Nara and Fiete}, and will serve as a reference solution in the following investigation.
However, the modal bandlimitation solution is available only for plane wave and point source virtual source models synthesized by a circular SSD, at the cost of increased computational complexity.

@@

\subsection{Spatial to temporal filter transform}
First, a general transformation is introduced allowing the transform of an arbitrary spatial filter to an equivalent temporal filter set.

@@

The shifted spatially filtered driving function reads as
\begin{multline}
    d'_x(s,t+ \tau(s) ) = \int h_x(s-s_0) \, d(s_0,t+ \tau(s) ) \, \td s_0 = \\
    \int h_x(s-s_0) \, A^D(s_0)\, \delta(s_0,t+ \left(\tau(s)-\tau(s_0)\right)  ) \, \td s_0.
    \label{eq:spat_conv2}
\end{multline}
In order to evaluate the spatial convolution the generalized sifting property of the Dirac-delta may be applied \cite{Firtha2019phd}, which states that
\begin{equation}
    \int f(s_0) \delta(g(s_0)) \td s_0 = \sum_i \frac{f(s_i)}{| \frac{\partial}{\partial s_0} g(s_0) |_{s = s_i}}, \hspace{3mm} g(s_i) = 0.
\end{equation}
In the present problem the zeros of the Dirac's argument are found where
\begin{equation}
    \tau(s_i) = t + \tau(s)
    \label{eq:zero}
\end{equation}
is satisfied.
In the following it is assumed that a single zero $s_i$ exists, satisfying \eqref{eq:zero}.
This assumption means that the wavefront arrives at each SSD element at a unique time instant, being strictly true for a virtual plane wave and a linear SSD.
The derivative of the phase function is given by
\begin{equation}
    \frac{\partial}{\partial s_0} g(s_0) = - \frac{\partial}{\partial s_0} \tau(s_0) = -\frac{\hat{k}_t^P(s_0)}{c},
\end{equation}
i.e. it is the tangential component of the local propagation vector.
With the above assumptions the integral \eqref{eq:spat_conv2} can be evaluated, yielding
\begin{equation}
    d'_x(s,t+ \tau(s) ) = h_x(s-s_i) \, A^D(s_i)
    .
    \label{eq:spat_conv_eval}
\end{equation}

@@

\subsubsection{Local plane wave approximation}
The above formulation already allows one to transform an arbitrary spatial impulse response into an equivalent temporal filter for each SSD element, as long as the virtual field model is known and \eqref{eq:zero} can be solved.
In order to give a more general solution it is assumed that the virtual field is locally plane, being a usual high frequency assumption in WFS theory.
As a further general WFS assumption it is supposed that the SSD is locally plane.
These requirements inherently ensure that a single solution exists for \eqref{eq:zero}\footnote{Except for a plane wave arriving normally to the SSD, at which case the temporal filter is transformed into a Dirac-delta.}.
With these assumptions the phase function is given as
\begin{equation}
    \tau(\vx) = \hat{k}^P_x(\vx)x +  \hat{k}^P_y(\vx) y
\end{equation}
and \eqref{eq:zero} is satisfied where
\begin{equation}
    s_i = s + \frac{c \cdot t}{\hat{k}_t^P(s)}
    .
\end{equation}
Finally, as a crucial approximation it is assumed that both the amplitude of the driving function and the propagation vector varies slowly along the SSD, i.e $A^D(s_i) \approx A^D(s)$ and $\hat{k}_t^P(s_i) \approx \hat{k}_t^P(s)$ holds.
With these assumptions the corresponding filter transform reads as
\begin{equation}
    h_t\left(s,t\right) =\frac{ c }{|\hat{k}_t^P(s)|} h_x\left(-t \,\frac{c}{\hat{k}_t^P(s)}\right).
    \label{eq:temporal_transform}
\end{equation}

@@

The validity and the error analysis of the above filter transform was investigated in details in \cite{Firtha DAGA2023}.
In the following the direct application for Local WFS is discussed.

@@

From Figure \ref{Fig:aliasing} (d) it is obvious, that the overlapping of the sampled driving function spectra, and, therefore, the presence of aliasing waves can be avoided by spatially bandlimiting the driving functions to the Nyquist wavenumber prior to sampling.
Spatial bandlimitation can be easily performed by designing a suitable spatial antialiasing filter and transforming it into an equivalent temporal filter bank, based on the results of the previous section.
An important advantage of the above temporal filtering is that the driving functions may be evaluated in the discrete SSD positions, and a subsequent temporal antialiasing inherently ensures analytical spatial bandlimitation.

@@

\subsubsection{Symmetrical antialiasing}
As the simplest antialiasing strategy, the spectral overlapping is avoided by the spatial lowpass filtering of the driving functions to the Nyquist wavenumber, $k_{s,\mathrm{Nyq}} = \frac{\pi}{\Delta s}$.

@@

From the transform given by \eqref{eq:frequency_transform} the angular frequency response of the equivalent temporal filter bank is given by
\begin{equation}
    H_t(s,\omega) = \frac{ 1 }{ \sqrt{ 1 + \left( \frac{\omega}{c}\frac{\hat{k}_t^P(s)}{k_{s,\mathrm{Nyq}}}  \right)^{2N} } } =  \frac{ 1 }{ \sqrt{ 1 + \left( \frac{\omega}{\omega_c(s)}  \right)^{2N} } } .
\end{equation}
with
\begin{equation}
    \omega_c(s) = c \frac{k_{s,\mathrm{Nyq}}}{\hat{k}_t^P(s)} = c\frac{ \pi}{\Delta s} \frac{1}{\hat{k}_t^P(s)}.
    \label{eq:cutoff_fr}
\end{equation}
Therefore, the cut-off frequency of the equivalent low-pass filter on a given SSD element is directly determined by the tangential component of the local propagation vector.

The result of antialiasing filtering is depicted in Figure \ref{Fig:symm_antialiasing} for the synthesis scenario discussed in the foregoing.
In the present case the sampling arc length is given by $\Delta s = \frac{2 \pi R_0}{N}$, i.e. the space-dependent cut-off frequency reads as\begin{equation}
    \omega_c(s) = \frac{ N}{ 2 R_0 } \frac{c}{\hat{k}_t^P(s)}.
\end{equation}
The antialiasing filters were a 4-th order Butterworth deisgn.

Figure (a) shows the result of the synthesis, while Figure (b) illustrates the spectrum of the bandlimited, sampled driving functions.
\begin{figure}[h!]
    \begin{center}
        \begin{overpic}[width = 0.75\columnwidth]{figs/antialiased_synthesis.png}
            \footnotesize \put(0,2){(a)}
        \end{overpic}
        \\
        \begin{overpic}[width = 0.75\columnwidth]{figs/antialiased_synthesis_spectrum.png}
            \footnotesize \put(0,0){(b)}
        \end{overpic}
    \end{center}
    \caption{Symmetric antialiasing filtering}
    \label{Fig:symm_antialiasing}
\end{figure}
It is highlighted, that the presented antialiasing strategy ensures highly suppressed aliasing waves in the center of the circular source array.
In other positions of the listening area lateral aliasing wavefronts remain in the reproduced wave field, while the target, virtual wave front is bandlimited.
This phenomena can be easily understood by investigating the synthesis scenario in a geometrical manner:

>>>>

\author{\IEEEauthorblockN{Gergely Firtha}
    \IEEEauthorblockA{\textit{Dept. of Networked Systems and Services} \\
        \textit{Budapest University of Technologies and Economics}\\
        H-1111 Budapest, Hungary \\
        firtha@hit.bme.hu}
    \and
    \IEEEauthorblockN{Nara Hahn}
    \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
        \textit{name of organization (of Aff.)}\\
        City, Country \\
        email address or ORCID}
    \and
    \IEEEauthorblockN{Frank Schultz}
    \IEEEauthorblockA{\textit{Institute of Communications Engineering} \\
        \textit{University of Rostock}\\
        18059 Rostock, Germany \\
        frank.schultz@uni-rostock.de}
    \and
    \IEEEauthorblockN{Péter Fiala}
    \IEEEauthorblockA{\textit{Dept. of Networked Systems and Services} \\
        \textit{Budapest University of Technologies and Economics}\\
        H-1111 Budapest, Hungary \\
        fiala@hit.bme.hu}
}

@@

\section{Introduction}
The aim of sound field synthesis is to reproduce a virtual target sound field over an extended listening area using a densely spaced loudspeaker arrangement, known as the secondary source distribution (SSD).
By feeding the loudspeakers with specific driving functions, the superposition of sound fields from each SSD element should ideally match the target sound field in the intended receiving area.
One prominent sound field synthesis method is Wave Field Synthesis (WFS) \cite{Berkhout1993:Acoustic_control_by_WFS, Start1997:phd}.

As WFS obtains the necessary driving functions from a suitable boundary integral representation of the target sound field, perfect phase correct field-reconstruction could be only achieved by employing a theoretical, continuous secondary source distribution.
In practical applications the SSD constist of discrete loudspeaker elements, resulting in secondary wavefronts additionally to the intended virtual wavefront, emerging from the individual SSD elements, termed as aliasing waves.
This spatial aliasing phenomena is perceived as the coloration of the synthesized sound field above the spatial aliasing frequency, varying with the listening position and the virtual wavefront direction, heavily degrading the quality of reproduction.

The effect of aliasing can be mitigated by restricting the synthesis to a smaller control region inside the listening area at the cost of permitting stronger aliasing artifacts outside the control region.
These anti-aliased WFS approaches are commonly referred to as Local Wave Field Synthesis (LWFS).
Conventional LWFS approaches achieve increased local synthesis accuracy by the direct spatial bandlimitation of the WFS driving functions \cite{}.
However, due to the high computational demand the real-time application of these LWFS approaches is not straightforward.

The present contribution introduces a novel LWFS approach, relying merely on the temporal bandlimitation of the SSD driving functions.
The temporal filters, defined uniquely for each SSD elements, are obtained from spatial antialiasing filters by means of an analytical spatial-to-temporal filter transformation, introduced by the current authors in a previous contribution \cite{firtha_daga}.
Similarly to other LWFS techniques, the presented approach ensures aliasing-free synthesis at a predefined listening position at the cost of temporally bandlimited sound field at other listening regions.
The results of the proposed approach are compared with a conventional LWFS approach employing direct spatial bandlimitation.

\section{Theoretical basics}
\subsection{The local propagation vector:}
Consider an arbitrary steady-state free space sound field at an angular frequency $\omega$.
As a standard ansatz in the field of geometrical acoustics the sound field can be written in the general polar from as
%
\begin{equation}
    P(\vx,\omega) = A(\vx) \te^{- \ti \omega \tau(\vx)},
    \label{eq:Pxw}
\end{equation}
%
where $A(\vx)$ and $\tau(\vx)$ are real-valued functions.
This formulation applies to both plane waves and (3D) point sources.
The propagation dynamics of the sound field are governed by its phase function $\tau(\vx)$, termed as the eikonal in the field of ray acoustics.
In the temporal domain, the sound field can be obtained by taking the inverse Fourier transform of \eqref{eq:Pxw}, yielding:
\begin{equation}
    p(\vx,t) = A(\vx) \, \delta \! \left( t - \tau(\vx)\right).
\end{equation}
From the above formulation it is evident that the eikonal $\tau(\vx)$ describes the propagation delay the wavefront takes to arrive at $\vx$.
The eikonal equation is obtained by substituting the ansatz \eqref{eq:Pxw} into the Helmholtz equation, stating that in a source free volume
\begin{equation}
    \left| \nabla \tau(\vx) \right| = \frac{1}{c}
\end{equation}
is fulfilled.
The gradient of the eikonal is termed as the local propagation vector
\begin{equation}
    \hat{\vk}^P(\vx) = \posvec{3}{\hat{k}_x^P(\vx)}{\hat{k}_y^P(\vx)}{\hat{k}_z^P(\vx)} = c \nabla \tau(\vx,\omega).
\end{equation}
In steady-state, the local propagation vector is perpendicular to the wavefront (equiphase positions) at an arbitrary location of unit length pointing towards the local propagation direction \cite{Firtha2016}.
In the temporal domain it points into the local propagation direction at the time instant of the wavefront passby.

@@

\subsection{Conventional 2.5D Wave Field Synthesis}
Consider a smooth convex secondary source distribution located at $\vxo = \posvec{3}{x_0}{y_0}{0}$ consisting of a continuous distribution of 3D point sources, described by the 3D freefield Green's function.
In this geometry the target field inside the area bounded by the SSD is described by the  Kirchhoff approximation of 2.5D Kirchhoff-Helmholtz integral, from which the 2.5D driving functions can be extracted.
For an arbitrary simple target virtual sound field $P(\vxo,\omega)$ the steady-state driving functions read as \cite{Firtha2016}
\begin{equation}
    D(\vxo,\omega) = \underbrace{\sqrt{8\pi \frac{\ti \omega}{c}}}_{ H_{\text{pre}}(\omega) }
    w(\vxo) \sqrt{d_{\mathrm{ref}}(\vxo)}
    P(\vxo,\omega)
    .
    \label{Eq:WFS25D}
\end{equation}
The driving function consists of
\begin{itemize}
    \item a frequency dependent pre-equalization filter $H_{\text{pre}}(\omega)$,
    \item a secondary source selection window
          \begin{equation}
              w(\vxo) = \max{\left( \hat{k}_{\mathrm{n}}^P(\vxo),0\right)},
          \end{equation}
          where $\hat{k}_{\mathrm{n}}^P(\vxo)$ is the normal component of the local propagation vector,
    \item a gain factor $\sqrt{d_{\mathrm{ref}}(\vxo)}$ allowing amplitude correction along a reference curve, depending on the actual virtual field model (c.f. \cite{Firtha2016}),
    \item and the virtual field measured on the SSD.
\end{itemize}
Assuming a simple virtual sound field as given by \eqref{eq:Pxw} the driving functions can be written as
\begin{equation}
    D(\vxo,\omega) = H_{\text{pre}}(\omega)
    \underbrace{ w(\vxo)  \, \sqrt{d_{\mathrm{ref}}(\vxo)} \, A(\vxo) }_{A^D(\vxo)} \te^{- \ti \omega \tau(\vxo)}
    ,
\end{equation}
and in the time domain as
\begin{equation}
    d(\vxo,t) = h_{\text{pre}}(t) \ast_t  A^D(\vxo)\,  \delta\left( t - \tau(\vxo)\right),
    \label{eq:d_wfs_td}
\end{equation}
with $A^D(\vxo)$ being the real valued overall gain factor of the driving function, $\ast_t$ denoting temporal convolution and $h_{\text{pre}}(t)$ is the temporal WFS prefilter impulse response.
The analytical pre-equalization filter impulse response is given in \cite{Deregowski1983, Firtha2019phd}, while for an ideal FIR filter implementation the reader is referred to \cite{Schultz2016}.
In the following due to the associativity of convolution this prefiltering is excluded from the discussion.

@@

\begin{figure*}[h!]
    \begin{center}
        \begin{overpic}[width = 0.75\columnwidth]{figs/ideal_synthesis.png}
            \footnotesize \put(0,2){(a)}
        \end{overpic} \hspace{2cm}
        \begin{overpic}[width = 0.75\columnwidth]{figs/aliased_synthesis.png}
            \footnotesize \put(0,2){(b)}
        \end{overpic}
        \\
        \begin{overpic}[width = 0.75\columnwidth]{figs/ideal_synthesis_spectrum.png}
            \footnotesize \put(0,0){(c)}
        \end{overpic} \hspace{2cm}
        \begin{overpic}[width = 0.75\columnwidth]{figs/aliased_synthesis_spectrum.png}
            \footnotesize \put(0,0){(d)}
        \end{overpic}
    \end{center}
    \caption{The synthesized field and the spatial spectrum of the driving function in case of ideal, quasi-continuous SSD (a) and (c), and in case of a discretized SSD (b) and (d).
    It is important ot note that the spatial spectrum depicts only the non-zero Fourier components (i.e. the Fourier series coefficients) normalized to the sampling wavenumber.}
    \label{Fig:aliasing}
\end{figure*}
Conventional WFS theory assumes a continuous secondary source distribution.
In practical applications the SSD is composed of evenly spaced discrete source elements.
As a result, \emph{aliasing wavefronts} emerge from the individual secondary sources, following the intended virtual wavefront, leading to a \emph{spatial aliasing phenomena}.
Although localization does not degrade due to the precedence effect, aliasing is perceived as strong coloration predominantly in the high-frequency region varying with the receiver position and the virtual source position/direction, occurring predominantly above the \emph{aliasing frequency}.

Figure \ref{Fig:aliasing} illustrates spatial aliasing in the temporal domain through the example of the synthesis of an impulsive plane wave, arriving from $\phi_{\mathrm{PW}} = 0^{\circ}$, applying a discrete secondary source distribution in comparison with a quasi-continuous synthesis scenario.
In the current example the SSD is a circular one with the radius of $R_0 = 2~\mathrm{m}$, consisting of $N_0 = 90$ 3D point sources.
The aliased synthesis in the temporal domain is depicted in Figure \ref{Fig:aliasing} (b).

Mathematically, aliasing can be modeled as the discretization of the theoretically continuous driving functions:
Assume that a suitable parametrization $d(s,t)$ of the secondary source is given, e.g. the arc length $s = R_0\, \cos \phi$ in case of a circular SSD (where $\phi$ is the polar angle), or the linear position on an infinite long linear SSD.
The discretization process is given by sampling the driving function by a sampling function, consisting of a series of Dirac-deltas in the actual secondary source positions:
\begin{equation}
    d_s(s,t) = \sum_{\nu = 0}^{N_0-1} d(s,t) \, \delta(s - \nu \Delta s) \, \Delta s,
    \label{eq:sampling_td}
\end{equation}
where $\Delta s$ is the sampling arc length.

By investigating the sampling process in the frequency-wavenumber domain, the spectrum of the sampled driving function is given by
\begin{equation}
    \tilde{D}_s(k_s,\omega) = \sum_{\nu = -\infty}^{\infty} \tilde{D}\left(k_s- \nu \frac{2\pi}{\Delta s},\omega\right) \Delta s,
    \label{eq:sampling_fd}
\end{equation}
where $t \rightarrow \omega$ and $s \rightarrow k_s$ are Fourier-transform pairs.
Obviously, for a closed SSD contour (most notablz for a circular SSD) the continuous spatial Fourier-transform yields a discrete spectrum, consisting of non-zero components only on the multiples of the spatial wavenumber, i.e. at
\begin{equation}
    k_s = \frac{m}{R_0},
\end{equation}
constituting a Fourier series.

@@

According to the sampling theorem, spatial aliasing can be avoided by analytically spatial-bandlimiting the driving functions to the Nyquist wavenumber.
This can be performed by either spatially filtering the driving functions with an appropriate spatial low-pass filter prior to numerical evaluation, or performing bandlimitation directly in the wavenumber (modal) domain analytically.
The latter approach is implemented by \cite{Nara and Fiete}, and will serve as a reference solution in the following investigation.
However, the modal bandlimitation solution is available only for plane wave and point source virtual source models synthesized by a circular SSD, at the cost of increased computational complexity.

@@

\subsection{Spatial to temporal filter transform}
First, a general analytical approach is introduced allowing the transform of an arbitrary spatial filter to an equivalent temporal filter set.
From herein, subscript $x$ and $t$ distinguishes wether a filter is defined as a spatial or the temporal filters, respectively.

@@

The shifted spatially filtered driving function reads as
\begin{multline}
    d'_x(s,t+ \tau(s) ) = \int h_x(s-s_0) \, d(s_0,t+ \tau(s) ) \, \td s_0 = \\
    \int h_x(s-s_0) \, A^D(s_0)\, \delta(s_0,t+ \left(\tau(s)-\tau(s_0)\right)  ) \, \td s_0.
    \label{eq:spat_conv2}
\end{multline}
In order to evaluate the spatial convolution the generalized sifting property of the Dirac-delta may be applied \cite{Firtha2019phd}, which states that
\begin{equation}
    \int f(s_0) \delta(g(s_0)) \td s_0 = \sum_i \frac{f(s_i)}{| \frac{\partial}{\partial s_0} g(s_0) |_{s = s_i}}, \hspace{3mm} g(s_i) = 0.
\end{equation}
In the present problem the zeros of the Dirac's argument are found where
\begin{equation}
    \tau(s_i) = t + \tau(s)
    \label{eq:zero}
\end{equation}
is satisfied.
In the following it is assumed that a single zero $s_i$ exists, satisfying \eqref{eq:zero}.
This assumption means that the wavefront arrives at each SSD element at a unique time instant, being strictly true for a virtual plane wave and a linear SSD.
The derivative of the phase function is given by
\begin{equation}
    \frac{\partial}{\partial s_0} g(s_0) = - \frac{\partial}{\partial s_0} \tau(s_0) = -\frac{\hat{k}_t^P(s_0)}{c},
\end{equation}
i.e. it is the tangential component of the local propagation vector.
With the above assumptions the integral \eqref{eq:spat_conv2} can be evaluated, yielding
\begin{equation}
    d'_x(s,t+ \tau(s) ) =  \frac{ c }{|\hat{k}_t^P(s_i)|} \, h_x(s-s_i) \, A^D(s_i)
    .
    \label{eq:spat_conv_eval}
\end{equation}

@@

\subsubsection{Local plane wave approximation}
The above formulation already allows one to transform an arbitrary spatial impulse response into an equivalent temporal filter for each SSD element, as long as the virtual field model is known and \eqref{eq:zero} can be solved.
In order to give a more general solution it is assumed that the virtual field is locally plane, being a usual high frequency assumption in WFS theory.
As a further general WFS assumption it is supposed that the SSD is locally plane.
These requirements inherently ensure that a single solution exists for \eqref{eq:zero}\footnote{Except for a plane wave arriving normally to the SSD, at which case the temporal filter is transformed into a Dirac-delta.}.
With these assumptions the phase function is given as
\begin{equation}
    \tau(\vx) = \hat{k}^P_x(\vx) \, x +  \hat{k}^P_y(\vx) \, y
\end{equation}
and \eqref{eq:zero} is satisfied where
\begin{equation}
    s_i = s + \frac{c \cdot t}{\hat{k}_t^P(s)}
    .
\end{equation}
Finally, as a crucial approximation it is assumed that both the amplitude of the driving function and the propagation vector varies slowly along the SSD, i.e $A^D(s_i) \approx A^D(s)$ and $\hat{k}_t^P(s_i) \approx \hat{k}_t^P(s)$ holds.
With these assumptions the corresponding filter transform reads as
\begin{equation}
    h_t\left(s,t\right) =\frac{ c }{|\hat{k}_t^P(s)|} h_x\left(-t \,\frac{c}{\hat{k}_t^P(s)}\right).
    \label{eq:temporal_transform}
\end{equation}

@@

The validity and the error analysis of the above filter transform was investigated in details in \cite{Firtha DAGA2023}.
In the following the direct application for local WFS is discussed.

@@

Figure \ref{Fig:aliasing} (d) reflects that the overlapping of the sampled driving function spectra, and, therefore, the presence of aliasing waves can be avoided by spatially bandlimiting the driving functions to the Nyquist wavenumber prior to sampling.
Spatial bandlimitation can be straightforwardly performed by designing a suitable spatial antialiasing filter and transforming it into an equivalent temporal filter bank, based on the results of the previous section.
An important advantage of the above temporal filtering is that the driving functions may be evaluated in the discrete SSD positions, and a subsequent temporal antialiasing inherently ensures analytical spatial bandlimitation.

@@

\subsubsection{Symmetrical antialiasing}
As the most straightforward antialiasing strategy, the spectral overlapping is avoided by the spatial lowpass filtering of the driving functions to the Nyquist wavenumber, $k_{s,\mathrm{Nyq}} = \frac{\pi}{\Delta s}$.

@@

From the transform given by \eqref{eq:frequency_transform} the angular frequency response of the equivalent temporal filter bank is given by
\begin{equation}
    H_t(s,\omega) = \frac{ 1 }{ \sqrt{ 1 + \left( \frac{\omega}{c}\frac{\hat{k}_t^P(s)}{k_{s,\mathrm{Nyq}}}  \right)^{2N} } } =  \frac{ 1 }{ \sqrt{ 1 + \left( \frac{\omega}{\omega_c(s)}  \right)^{2N} } } .
\end{equation}
with
\begin{equation}
    \omega_c(s) = c \frac{k_{s,\mathrm{Nyq}}}{\hat{k}_t^P(s)} = c\frac{ \pi}{\Delta s} \frac{1}{\hat{k}_t^P(s)}.
    \label{eq:cutoff_fr}
\end{equation}
Therefore, the cut-off frequency of the equivalent low-pass filter on a given SSD element is inversely proportional to the tangential component of the local propagation vector.

The result of antialiasing filtering is depicted in Figure \ref{Fig:symm_antialiasing} for the synthesis scenario discussed in the foregoing.
In the present case the sampling arc length is given by $\Delta s = \frac{2 \pi R_0}{N}$, i.e. the space-dependent cut-off frequency reads as 
\begin{equation}
    \omega_c(s) = \frac{ N}{ 2 R_0 } \frac{c}{\hat{k}_t^P(s)}.
\end{equation}
The antialiasing filters were a 4-th order Butterworth deisgn.

Figure (a) shows the result of the synthesis, while Figure (b) illustrates the spectrum of the bandlimited, sampled driving functions.
\begin{figure}[h!]
    \begin{center}
        \begin{overpic}[width = 0.75\columnwidth]{figs/firtha_center.png}
            \footnotesize \put(0,2){(a)}
        \end{overpic}
        \\
        \begin{overpic}[width = 0.75\columnwidth]{figs/antialiased_synthesis_spectrum.png}
            \footnotesize \put(0,0){(b)}
        \end{overpic}
    \end{center}
    \caption{Symmetric antialiasing filtering. Figure (a) depicts the synthesized wave front, and (b) illustrates the spatial spectrum of the driving functions.}
    \label{Fig:symm_antialiasing}
\end{figure}
It is highlighted that the presented antialiasing strategy ensures suppressed aliasing waves in the center of the circular source array.
In other positions of the listening area lateral aliasing wavefronts remain in the reproduced wave field, while the target, virtual wave front is bandlimited.
This phenomena can be easily understood by investigating the synthesis scenario in a geometrical manner:
